The authors would like nothing more than to dive right into the sheer excitement of Precalculus. However, experience - our own as well as that of our colleagues - has taught us that is it beneficial, if not completely necessary, to review what students should know before embarking on a Precalculus adventure.  The goal of Chapter \ref{Prerequisites} is exactly that: to review the concepts, skills and vocabulary we believe are prerequisite to a rigorous, college-level Precalculus course.  This review is not designed to teach the material to students who have never seen it before thus the presentation is more succinct and the exercise sets are shorter than those usually found in an Intermediate Algebra text.  An outline of the chapter is given below.

\smallskip

Section \ref{SetTheory} (Basic Set Theory and Interval Notation) contains a brief summary of the set theory terminology used throughout the text including sets of real numbers and interval notation.

\smallskip

Section \ref{RealNumberArithmetic} (Real Number Arithmetic) lists the properties of real number arithmetic.\footnote{You know, the stuff students mess up all of the time like fractions and negative signs.  The collection is close to exhaustive and definitely exhausting!}

\smallskip

Section \ref{LinearEqIneq} (Linear Equations and Inequalities) focuses on solving linear equations and linear inequalities from a strictly algebraic perspective.  The geometry of graphing lines in the plane is deferred until Section \ref{LinearFunctions} (Linear Functions).

\smallskip

Section \ref{AbsValEqIneq} (Absolute Value Equations and Inequalities) begins with a definition of absolute value as a distance.  Fundamental properties of absolute value are listed and then basic equations and inequalities involving absolute value are solved using the `distance definition' and those properties.  Absolute value is revisited in much greater depth in Section \ref{AbsoluteValueFunctions} (Absolute Value Functions).

\smallskip

Section \ref{PolyArith} (Polynomial Arithmetic) covers the addition, subtraction, multiplication and division of polynomials as well as the vocabulary which is used extensively when the graphs of polynomials are studied in Chapter \ref{Polynomials} (Polynomials).

\smallskip

Section \ref{Factoring} (Factoring) covers basic factoring techniques and how to solve equations using those techniques along with the Zero Product Property of Real Numbers.

\smallskip

Section \ref{QuadEqus} (Quadratic Equations) discusses solving quadratic equations using the technique of `completing the square' and by using the Quadratic Formula.  Equations which are `quadratic in form' are also discussed.

\smallskip

Section \ref{RatExpEqus} (Rational Expressions and Equations) starts with the basic arithmetic of rational expressions and the simplifying of compound fractions.  Solving equations by clearing denominators and the handling negative integer exponents are presented but the graphing of rational functions is deferred until Chapter \ref{Rationals}  (Rational Functions).

\smallskip

Section \ref{RadEqus} (Radicals and Equations) covers simplifying radicals as well as the solving of basic equations involving radicals.  

\smallskip

Section \ref{CmpNums} (Complex Numbers) covers the basic arithmetic of complex numbers and the solving of quadratic equations with complex solutions.

\newpage

\section{Basic Set Theory and Interval Notation}

\input{SetTheory}

\newpage

\section{Real Number Arithmetic}

\input{RealNumberArithmetic}

\newpage

\section{Linear Equations and Inequalities}

\input{LinearEqIneq}

\newpage

\section{Absolute Value Equations and Inequalities}

\input{AbsValEqIneq}

\newpage

\section{Polynomial Arithmetic}

\input{PolyArith}

\newpage

\section{Factoring}

\input{Factoring}

\newpage

\section{Quadratic Equations}

\input{QuadEqus}

\newpage

\section{Rational Expressions and Equations}

\input{RatExpEqus}

\newpage

\section{Radicals and Equations}

\input{RadEqus}

\newpage

\section{Complex Numbers}

\input{CmpNums}

\newpage